---
title: "Afternoon Practice"
author: "Clarissa Boyajian"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Warm-up:
Look at `us_rent_income` (built into the `tidyr` package). Make a tidy version. Then from your tidy version, get it back to the original structure.

```{r}
us_rent_income_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)

#us_rent_income_long <- us_rent_income_wide %>% 
  #pivot_longer(cols = income:rent,
               #names_to = variable,
               #values_to = estimate)

```

## Practice soem tidyverse functions for data wrangling:
- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- pipiing things together in sequence



## dplyr::filter() to keep or exclude ROWS based on our conditions

#### 1. exact match (single condition)

Only keep observations where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
```

#### 2. "and statement within filter

Only keep observations if the penguin species is Chinstrap AND the year is 2008
```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap" & year == 2008)
  #can also write it like this: `filter(species == "Chinstrap", year == 2008)`
  #or like this `filter(species == "Chinstrap") %>% filter(year == 2008)`
```

#### 3. Exclude Gentoos

Keep observations if the penguin species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

#### Exercise: 

Create a subset from penguins that only contains observations from Adelie penguins with a flipper length greater than or equal to 185 mm. 

```{r}
practice_1 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)
```

#### 4. Using the %in% operator

Use the %in% operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row

Keep observations where the island is Dream or Torgersen
```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
```
Neighbor chat: what would actually be happening in the code above if I'd used `==` instead of `%in%`
- `==` will rotate through the vector in order and look for a match for each value in each row (not all values in all rows)

Keep observations where the island value is NOT Dream or Torgersen
```{r}
ex_4_1 <- penguins %>% 
  filter(!island %in% c("Dream", "Torgersen"))
```



## `select()` to select or exclude COLUMNS

#### 5. Keep a single column by name

Only want to keep the species column
```{r}
ex_5 <- penguins %>% 
  select(species)

# Reminder: re-write this without pipe operator
ex_5_1 <- select(.data = penguins, species)
```

#### 6. Let's keep multiple columns by name

Create a subset that contains species, bill_lenth_mm through flipper_lenth_mm, and year
```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

#### 7. Choose or exclude columns based on position

Create subset that excludes the 4th column (bill_depth_mm)
```{r}
ex_7 <- penguins %>% 
  select(-4)
```

#### -- practice -- 8. A piped sequence

Write a piped sequence where you first: only keep columns for species, body_mass_g, and island, and second: filter to only include observations if body mass is greater than 3600
```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```



## `mutate()` to add or update existing columns, while keeping the ones already there (OR transform existing column somehow)

#### 9. Add a column that has body mass converted to kg

If you tell it a column name that already exists, it will overwrite the existing column with new info after `=`
If you type a new name, it will create a new column with new info after `=`

*good idea to create new column always (not write over existing columns). Can always select out old columns later*
```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm / 1000)
  #can also write as `mutate(body_mass_kg = body_mass_g / 1000, flipper_length_m = flipper_length_mm / 1000)
```

#### 10. Update existing column 
(use to replace common spelling throughout data set)

If ever in doubt, create NEW column

```{r}
ex_10 <- penguins %>% 
  mutate(island_hooray = str_replace(island, pattern = "sen", replacement = "HOORAY"))
```

#### -- practice -- 11. Update the class of the bill depth column to character

```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))
```



## `case_when()` - your favorite if-else

#### 12. Create a new column that contains "small" if penguin flipper length is less than 185 mm, and "medium" otherwise

When creating new column have to do it within `mutate()`
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```

#### Task: starting from penguins, create new column called "bird_bill_bin" that contains:
  - if bill length is less than 38 mm: "short bill"
  - if bill length greater than or equal to 38mm and less than 43mm: "medium bill"
  - if bill length greater than or equal to 43mm: "long bill
  
*good idea to make if-else statements as explicit as possible*
```{r}
practice_2 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
    #can have final condition be anything else, write as: `TRUE ~ "long bill"` -- means NA values would be included too -- HAS to      be last statement to be catch-all
  ))
```



## `group_by() %>% summarize()` to find summary values by group and return them in a nice summary table

#### 13. Find mean of bill depth by species type

Creates summary table with two new columns, "mean_flipper_length" and "sd_flipper_length", that shows this grouped by species
```{r}
ex_13 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_filipper_length = sd(flipper_length_mm, na.rm = TRUE))

#Not the way to do it (probably) -- won't give you a summary table, will add a new column that lists mean for each species on each row based on species
ex_13_1 <- penguins %>% 
  group_by(species) %>% 
  mutate(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
```

#### Practice: 
In a single pipe sequence:
- only keep penguins with body mass greater than 3500g
- only keep columns for species, flipper length, and sex
- create a summary table that contains the mean, sd, max, and min value of flipper length grouped by species and sex

```{r}
practice_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flip = mean(flipper_length_mm, na.rm = TRUE),
            sd_flip = sd(flipper_length_mm, na.rm = TRUE),
            min_flip = min(flipper_length_mm, na.rm = TRUE),
            max_flip = max(flipper_length_mm, na.rm = TRUE)) %>% 
  ungroup() %>% 
  #deals with un-grouping iin summary table (useful if you're continuing to use the summary table in analysis)
  drop_na(sex)
  #will exclude all data from any penguins where sex wasn't recorded
```



