---
title: "TIdying Data Practice"
author: "Clarissa Boyajian"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

## Read in the wb-indicators.csv file, make sure to specify 'na = c("..", "")`
```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..", ""))

```

## Tidying Data
- Could have all of this done as one long pipe sequence (rule of thumb is 7-8 in one pipe sequence). Doing it this way to be able to look at each step. 

#### Reshape so that year is a single column
```{r}
wb_indicators_longer <- wb_indicators %>% 
  pivot_longer(cols = `2001 [YR2001]`:`2020 [YR2020]`,
               names_to = "year",
               values_to = "indicator_value")
```

#### Clean up that year column (and drop unnecessary columns)
```{r}
wb_data_sept <- wb_indicators_longer %>% 
  separate(col = "year", 
           into = c("year", "year_chr"),
           sep = " ") %>% 
  select(-c(year_chr, `Country Code`, `Series Code`))
```

#### Use `pivot_wider()` to get all indicators in their own column (use `drop_na()` to not include rows that have no info in the "indicators_value" column)
```{r}
wb_data_tidy <- wb_data_sept %>% 
  drop_na(`Series Name`) %>% 
  pivot_wider(names_from = `Series Name`,
              values_from = indicator_value)
```

#### Clean up column names (rename ALL column names)
```{r}
names(wb_data_tidy) <- c("country", "year", "access_clean_fuels", "access_electricity", "c02_emissions", "fossil_fuel_cons", "water_stress")


# OR rename a single or some columns (NOTE: `rename(new name = old name)`)
wb_data_tidy <- wb_data_tidy %>% 
  rename(country_name = country,
         electricity = access_electricity)
```


## Practicing what we learned above
```{r}
df <- tribble(
  ~name, ~A, ~B, ~C,
  "teddy", 4, 5, 6,
  "khor", 9, 10 , 11
)

df

# If A, B, and C are different dog food types - convert into tidy structure

df_long <- df %>% 
  pivot_longer(col = A:C,
               names_to = "dog_food",
               values_to = "cans_per_day")


df_wider <- df_long %>% 
  pivot_wider(names_from = dog_food,
              values_from = cans_per_day)
```

